\chapter*{Acknowledgements} 
\addcontentsline{toc}{chapter}{Acknowledgements} 

This work was supported by funding from the Australian Research Council and the Australian Antarctic Division science program.
Funding for sequencing described in \secreft{ch:polarfront} was provided by the Gordon and Betty Moore Foundation to the J.\ Craig Venter Institute.

Many thanks to Captain Murray Doyle as well as the crew, science support team and most importantly the caterers of the RSV \textit{Aurora Australis} for keeping us afloat, productive and well-fed through two long voyages.
Special thanks are due to Steve Rintoul for his voyage leadership and invaluable advice.
Erik van Sebille provided crucial expertise and assistance on circulation modelling.

To Rick Cavicchioli and Federico Lauro, this PhD would not have been possible without your unfailing guidance.
To everyone in the RC group, past and present, I owe thanks for five years of support, friendship and sound advice.
Without Timothy Williams' seemingly boundless knowledge of biochemistry and microbial ecology I would have been lost at sea, and without his unparalleled good company aboard the \textit{Aurora} I would have been bored at sea.
Sheree Yau is the perceptive and indefatigable scientist the rest of us can only aspire to be, and is responsible for shooting down many of my worst ideas and refining many of the good ones.

Finally, thanks to James, Sarah, Alf, my housemates from the ``Casa de Zet'' and my family for your friendship and encouragement throughout my PhD.
